Automatic Matching of Bullet Lands

### 2016-04-19 revision/response.Rmd

7. There should have been a comparison of different classification methods to show that the random forest is reasonable.

***I am not quite sure what a 'reasonable' classification method is supposed to mean. Shouldn't a classification method be judged by its success in predicting the outcome? 
Apart from that, random forests are based on decision trees, which are an extremely flexible tool for classification, that can deal with quantitative and categorical types of variables and, in the implementation we used, also deal with some missing values (which was not needed in the data here).***

8. On the random forest classifier- are the prior probabilities being estimated from the cross comparisons described on page 15? What exactly are the prior probabilities used in the random forest?

***No data-based adjustment was made to the prior probabilities, i.e. prior probabilities of 50:50 for matches/non-matches were used.***

***HH: what does the reviewer mean by cross-comparison? On page 15 we discuss some results from matching based on CMS.***

@@

Reviewer 1:

21. First, the term "land" is used in the title but not defined until p. 3.  I recommend changing the title, to ensure broad comprehenson; perhaps "Automatic Matching of Bullet Marks," or something of that kind.

***HH: Let's discuss that on Monday***

***EH: Quick definition of land in abstract***

***HH: this is clearly a misunderstanding, so we need to be nice about breaking this to the reviewer, and emphasize it in the paper some more.***

22. Early on in the paper, it might be helpful to mention the typical width of a striation, the number of striations typically seen on a fired bullet, and so forth. 
***EH: Add typical width of groove + lands***

@@

23. p. 2, l. -1:  It would be good to reword so the sentence doesn't start with "x3p".  Also p. 3, l. 2.

***DONE***

24. p. 3, Fig. 1:  The numbers on the y-axis are superimposed.

***EH: We could slightly rotate the image in RGL, but the top-down view is nice. Alternatively I could just disable the axis labels - We don't really need to see them in this image, right?***

***HH: I just deleted the values from the axis***

@@

30. p. 13, l. -3:  You write "Given that striation patterns are typically much larger, ..." and I think it would be more clear to write "Given that striation widths are typically much larger, ...".

***DONE***

31. p. 16, l. -1/-2:  Could you elaborate on why a "large number of maximal CMS by itself is not indicative of a match"?  I think the reason is that there could be other marks that are completely incompatible and preclude a match, based on the next sentence, but I am not sure and a few words would be helpful.

***EH: From our data out of non matches its a 2 / 10212 event... but with a lot of pairwise comparisons, it will happen by chance even if its very unlikely. Adding a couple more bullets / comparisons we may see higher values.***

@@

34. p. 18, l. -4:  Change "this study" to "that study", so the reader is certain you mean Ma et al. (2001) and not your own.

***DONE***

35. p. 20,  l. -3:  I really like the idea of including an inconclusive zone.

***EH: Add thank you, and explain the ones that could be considered inconclusive. Lots more discussion would be needed about the extent of that zone.***

@@

41. So, my suggestion here is to change the sentence in the abstract to something like ``Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel." Or ``...help the trier of fact determine...".

***DONE***

42. the concept of ``match": The primary goal in forensic science is NOT to attempt to match two bullets. It is to quantify the value of the observations made on the two bullets with regard to the two competing propositions of interest to the court (i.e., assign a likelihood ratio or Bayes factor).

***EH: Don't call it primary goal, call it one of the primary goals***

@@

***DONE***

***HH: would a removal of the comma make it a complete sentence?***

***EH: I actually thought it already was a complete sentence... but I changed it to "The correlation peaks at a lag of -17, which indicates the largest amount of agreement between the signatures."***

***HH: Ta, as they say around here :)***

***EH: Change to the correlation is maximizes.***

### 2016-04-20 revision/response.Rmd

41. So, my suggestion here is to change the sentence in the abstract to something like ``Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel." Or ``...help the trier of fact determine...".

***DONE***

42. the concept of ``match": The primary goal in forensic science is NOT to attempt to match two bullets. It is to quantify the value of the observations made on the two bullets with regard to the two competing propositions of interest to the court (i.e., assign a likelihood ratio or Bayes factor).

***We have clarified that matching two bullets is one of the goals, but not the sole or primary goal.***

@@

43. I strongly encourage the authors to not follow this trend and use the term ``ballistics" only when you are actually talking about ballistics.

***We have removed references to the term ``ballistics" except in the context of citing prior work which uses the term.***

@@

48. Line 246: Replace ``this representations" with ``this representation" or ``these representations"

***DONE***

49. Caption of Figure 8: ``At a lag of -17 the correlation peaks, indicating the largest amount of agreement between the signatures." is not a complete sentence.

***We have revised this caption to remove the ambiguity from the word ``peaks", which we had intended to mean "maximizes."***

### 2016-04-20 imaging-paper.Rnw
added ccf to the list of features - where was that before??
Big set of changes

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is the primary goal. We propose an automated framework for the analysis of the 3D surface measurements, that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands with too much damage to be suitable for comparison, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} by a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} of the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify joint intervals between two signatures as those areas, in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features could be extracted. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

@@

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 8, fig.height = 3, out.width = '.65\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation function between the two signatures shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. The correlation peaks at a lag of -17, which indicates the largest amount of agreement between the signatures. Figure~\ref{fig:crosscutX} shows the lag-shifted signatures.}
\end{figure}

@@

We have presented an algorithm which detects the most prominent but least relevant structure of a bullet from a ballistics identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are matches. We created a random forest model using many of the features to provide a probabilistic assessment of the quality of a match, along with the most relevant features. %Finally, we have provided an easy-to-use open-source backend, and web frontend to many of these features.

@@

An approach that might be helpful comes from Biometrics \citep{jain:2015:tphil, jain:2015}. It has been successfully applied in the Aadhar project (the effort by the Unique Identification Authority of India (UIDAI) to provide a unique 12-digit identification number to approximately 1.2 billion residents of India based on ten fingerprints and two irises).
This approach uses dictionary learning based on small square patches  of 64x64 or 32x32 pixels. For matching bullets, a simpler, one-dimensional window approach might be sufficient to match striation marks because of the physical forces  leading to the existence of these marks.

\section*{Acknowledgment} 
Thanks to David Baldwin for pointing us to the NIST database and doing a Ballistics 101 for us.
Thanks to the men and women behind the software R~\citep{R}, and the authors of the R packages  knitr~\citep{knitr} and ggplot2~\citep{ggplot2}.

>>>>

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is one of the primary goals. We propose an automated framework for the analysis of the 3D surface measurements, that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands with too much damage to be suitable for comparison, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of convictions in the United States criminal justice system. Firearms identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of firearms identification evidence~\citep{giannelli:2011}.

@@

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} by a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} of the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify joint intervals between two signatures as those areas, in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features could be extracted. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item \hh{The value of the cross-correlation function (ccf) between the aligned signatures \cite{vorburger:2011}, }
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

@@

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 8, fig.height = 3, out.width = '.65\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation function between the two signatures shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. The correlation is maximized at a lag of -17, indicating the largest amount of agreement between the signatures. Figure~\ref{fig:crosscutX} shows the lag-shifted signatures.}
\end{figure}

@@

We have presented an algorithm which detects the most prominent but least relevant structure of a bullet from a firearms identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are matches. We created a random forest model using many of the features to provide a probabilistic assessment of the quality of a match, along with the most relevant features. %Finally, we have provided an easy-to-use open-source backend, and web frontend to many of these features.

@@

%An approach that might be helpful comes from Biometrics \citep{jain:2015:tphil, jain:2015}. It has been successfully applied in the Aadhar project (the effort by the Unique Identification Authority of India (UIDAI) to provide a unique 12-digit identification number to approximately 1.2 billion residents of India based on ten fingerprints and two irises).
%This approach uses dictionary learning based on small square patches  of 64x64 or 32x32 pixels. For matching bullets, a simpler, one-dimensional window approach might be sufficient to match striation marks because of the physical forces  leading to the existence of these marks.

\section*{Acknowledgment} 
Thanks to David Baldwin for pointing us to the NIST database and doing a Firearms 101 for us.
Thanks to the men and women behind the software R~\citep{R}, and the authors of the R packages  knitr~\citep{knitr} and ggplot2~\citep{ggplot2}.
